Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-02-10 manuscript/cpcrw_manuscript.Rmd
Misc changes
Changed WC presentation; named code chunks; cleaned up exclude/outlier
computation and text.

```{r, echo=FALSE, results='hide', message=FALSE, fig.keep='none'}
pclean <- function(x) {  # pretty formatting of p-values
  ifelse(x < 0.001, "< 0.001", paste("=", formatC(x, digits = 3, format = 'f'))) %>%
    paste("P", .)
}

library(dplyr, warn.conflicts = FALSE)
setwd("../")
source("R/5-analyze.R")
```

@@

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These groups included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status closely matched those of the 4 °C "drought" group. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

@@

```{r, echo=FALSE}
# TODO: this isn't good--fragile if files move or change names
# Want everything in-memory (or something)
# Compute outlier and exclusion numbers
sdn <- readLines("../outputs/summarydata.csv") %>% length
sdcn <- readLines("../outputs/summarydata_clean.csv") %>% length
en <- readLines("../outputs/3-remove/excluded_data.csv") %>% length
rn <- readLines("../outputs/3-remove/removed_data.csv") %>% length
```

Anomalous data were excluded based on their being more than 3.0 (for CO~2~) or 5.0 (for CH~4~) mean absolute deviations {Davies, 1993 #3138} from the treatment mean within each day, treatment, and temperature. We excluded `r en` of `r sdn` (`r round(en/sdn*100)`%) of the summarized samples for this reason. Other data (`r rn` of `r sdn`, `r round(rn/sdn*100)`%) were removed because of known instrument problems, e.g. the analyzer was left running after leaving a chamber. The final number of valid flux samples from the 100-day incubation was `r sdcn`.

@@

* Gas fluxes over incubation (Figures 2 and 3)
* Statistical tests of WC, T, C, N on individual core fluxes (Table 2)

@@

Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and field moisture cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C drought and controlled drought treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

* CO2:CH4 ratios
* Statistical tests of WC and temperature on cumulative fluxes

@@

```{r kable, echo=FALSE}
# Print out the `WC_effect` data frame in a nice-looking table
WC_effect$Gas[WC_effect$Gas == "CO2"] <- "CO<sub>2</sub>"  # html
WC_effect$Gas[WC_effect$Gas == "CH4"] <- "CH<sub>4</sub>"  # html
knitr::kable(WC_effect, digits = 3, col.names = c("Gas", "Treatment", "Temperature", "R<sup>2</sup>", "Sigma", "P", "AIC"))
```

@@

**Figure 1.** Core water content across the course of the experiment.

```{r, echo=FALSE}
print(figureA)
```

@@

```{r, echo=FALSE}
print(figureB)
```

@@

```{r, echo=FALSE}
print(figureC)
```

@@

```{r, echo=FALSE, warning=FALSE}
print(figureD)   # TODO: wtf is going on here?
```

>>>>

```{r setup, echo=FALSE, fig.keep='none', message=FALSE, results='hide'}
library(dplyr, warn.conflicts = FALSE)

pclean <- function(x) {  # pretty formatting of p-values
  ifelse(x < 0.001, "< 0.001", paste("=", formatC(x, digits = 3, format = 'f'))) %>%
    paste("P", .)
}

# Run the analysis script, which will leave a lot of figures, tables, and data in memory for us
setwd("../")
source("R/5-analyze.R")
```

@@

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These groups included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close matched those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

@@

```{r compute_excludes, echo=FALSE}
# Compute outlier and exclusion numbers
# TODO: this isn't good--fragile if files move or change names
# Want everything in-memory (or something)
rn <- readLines("../outputs/3-remove/removed_data.csv") %>% length
sdn <- readLines("../outputs/summarydata.csv") %>% length
fluxn <- nrow(fluxdata)
en <- sum(fluxdata$outlier)
```

Anomalous data were excluded based on their being more than `r CO2_EXCLUDE_DEVS` (for CO~2~) or `r CH4_EXCLUDE_DEVS` (for CH~4~) mean absolute deviations {Davies, 1993 #3138} from the treatment mean within each day, treatment, and temperature. We excluded `r en` of `r fluxn` (`r round(en/fluxn*100)`%) of the gas flux measurements for this reason. Other data (`r round(rn/sdn*100)`%) were removed because of known instrument problems, e.g. the analyzer was left running after leaving a chamber. The final number of valid flux samples from the 100-day incubation was `r sum(!fluxdata$outlier)`.

@@

* Gas fluxes over incubation (Figures 2 and 3)

* Statistical tests of WC, T, C, N on individual core fluxes (Table 2)

@@

Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

* CO2:CH4 ratios

```{r compute_ratios, echo=FALSE}
# Ratio data is stored in the `gas_ratio` d.f. computed by analysis script
minmag <- min(gas_ratio$CO2CH4_ratio) %>% log10 %>% floor %>%
  c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")[.]
minratio <- min(gas_ratio$CO2CH4_ratio / 1e6)  %>% formatC(digits = 1, format = 'f')
mintrt <- gas_ratio$Treatment[which.min(gas_ratio$CO2CH4_ratio)]
mintemp <- gas_ratio$Temperature[which.min(gas_ratio$CO2CH4_ratio)]
maxratio <- max(gas_ratio$CO2CH4_ratio / 1e6) %>% formatC(digits = 1, format = 'f')
maxtrt <- gas_ratio$Treatment[which.max(gas_ratio$CO2CH4_ratio)]
maxtemp <- gas_ratio$Temperature[which.max(gas_ratio$CO2CH4_ratio)]
```

The production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment.

@@

```{r table1, echo=FALSE}
# Print out the `WC_effect` data frame in a nice-looking table
WC_effect$Gas[WC_effect$Gas == "CO2"] <- "CO<sub>2</sub>"  # html
WC_effect$Gas[WC_effect$Gas == "CH4"] <- "CH<sub>4</sub>"  # html
knitr::kable(WC_effect, digits = 3, col.names = c("Gas", "Treatment", "Temperature", "R<sup>2</sup>", "Sigma", "P", "AIC"))
```

@@

**Figure 1.** Core water content across the course of the incubation experiment by temperature (left panel 4 °C, right panel 20 °C) and treatment.

```{r figureA, echo=FALSE}
figureA
```

@@

```{r figureB, echo=FALSE}
figureB
```

@@

```{r figureC, echo=FALSE}
figureC
```

@@

```{r figureD, echo=FALSE, warning=FALSE}
figureD
```
